Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
fix up various obvious things in section 5

@@

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible then to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given back by Reynolds
and Plotkin (cite) is CPS, which was specifically introduced by them
to make the evaluation strategy of the definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
  Reynolds that typed setting would be nice?}. Specifically, we
demonstrate a CBN evaluation of our object language in OCaml, which is
CBV. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by Plotkin (cite). As in the
more familiar CBV CPS, the interpretation of a value is a function
accepting the continuation argument |k| and yielding the answer
returned by that continuation.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = 
  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
\end{code}
In the both cases, the returned value has the type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

The interpretation of abstraction and application in CBN CPS is
different from CBV CPS and is as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
\end{code}
In the tell-tale sign of CBN, in interpreting the application
we do not `evaluate' the argument e2, that is, we do apply it to the
continuation. Rather, we pass the unevaluated |e2| to the abstraction.
The result of |lam (fun x -> add x (int 1))| has the type
|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) -> 'w2) -> 'w2|.
We want to collect the above interpretation functions into a module
with the Symantics signature, so to fit the CBN CPS interpreter within our
general framework. Alas, we encounter the same problem we had in
Section~\ref{S:PE-problem}, in writing a type equation for the |repr|
type that can represent both the type of |int i| and the type of 
|lam f|. We cannot say that the the result of interpreting an object
expression of the type |t| is |(t->'w)->'w| because if |t| happens to
be an arrow type such as |int->int| in the above example, the result
type obviously does not fit the desired pattern. We need a type
function with a type-case distinction. In Section~\ref{S:PE-solution} we have
shown the way to encode such a function, using the additional type
argument |'sv| of the |repr| type. Furthermore, we notice that the type 
function |repr_pe| needed for the partial evaluator,
Section~\ref{S:PE-solution}, is precisely the same type function we
need for CBN CPS. This gives us for the following CBN CPS interpreter,
of the Symantics signature:

@@

We have made our interpreter fully polymorphic over the answer type.
We could have just as well made |RCN| a functor, parameterized over
the answer type -- as is common when writing CPS in SML. In the
present case, we chose to use higher-rank polymorphism of OCaml record
types.

Because |RCN| has the Semantics signature, we can instnatiate with it and run
all of our previous examples. A more interesting is the following
example.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun self -> self)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
Section~\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ();;
\end{code}
we never get any result because of looping. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

\oleg{mention some formal results? Like indifference theorems by
  Plotkin? And say we generalized their result to the type setting?
  Mention this in the conclusions?}

@@

In this section however we show a more general approach to CBV CPS:
CPS transformer. We take any implementation of Symantics (interpreter,
compiler, partial evaluator) and make it evaluate the CPS language
instead. We can view this transformation of interpreters as a
transformation on the object language, from direct to
continuation-passing style.

@@

\section{Lifting and Functor}

>>>>

\jacques{we should probably be more explicit about the contributions of
the discussion section}

@@

We show that our |Symantics| and generally our approach accomodates
a number of variants without any (new) difficulties.  In particular,
we show how a a call-by-name CPS interpreter, a call-by-value CPS
program transformation, as well as how to deal with imperative features.
Finally, we comment on the requirements on the type system to support 
our approach.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\cite{ReynoldsPlotkin} is continuation-passing style (CPS), which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
Reynolds that typed setting would be nice?}\jacques{I think so, but from
which paper? The one ReynoldsPlotkin paper is typed}. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by Plotkin \cite{PlotkinCBN}. As in
the more familiar CBV CPS, the interpretation of a value is a function
accepting a continuation argument |k| and yielding the answer
returned by that continuation.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = 
  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

The interpretation of abstraction and application in CBN CPS is
different from CBV CPS and is as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
\end{code}
In the tell-tale sign of CBN, in interpreting the application
we do not `evaluate' the argument e2, that is, we do apply it to the
continuation. Rather, we pass the unevaluated |e2| to the abstraction.
The result of |lam (fun x -> add x (int 1))| has the type
|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) -> 'w2) -> 'w2|.
We want to collect the above interpretation functions into a module
with the |Symantics| signature, so as to fit the CBN CPS interpreter within our
general framework. Alas, we encounter the same problem we had in
Section~\ref{S:PE-problem}, in writing a type equation for the |repr|
type that can represent both the type of |int i| and the type of 
|lam f|. We cannot say that the result of interpreting an object
expression of type |t| is |(t->'w)->'w| because if |t| happens to
be an arrow type (such as |int->int| in the above example), the result
type obviously does not fit the desired pattern. We need a type
function with a typecase distinction. In Section~\ref{S:PE-solution} we have
shown how to encode such a function, using the extra type
argument |'sv| of the |repr| type. Furthermore, we notice that the type 
function |repr_pe| needed for the partial evaluator 
(in~\S\ref{S:PE-solution}) is precisely the same type function we
need for CBN CPS. This gives us for the following CBN CPS interpreter,
with the |Symantics| signature:

@@

We have made our interpreter fully polymorphic over the answer type.
We could have just as easily made |RCN| a functor, parameterized over
the answer type -- as is common when writing CPS in SML. In the
present case, we chose to use the higher-rank polymorphism of offered by
OCaml record types.

Because |RCN| has the signature Symantics, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the following example.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun self -> self)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
Section~\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ();;
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

\oleg{mention some formal results? Like indifference theorems by
  Plotkin? And say we generalized their result to the type setting?
  Mention this in the conclusions?}
\jacques{you have strayed outside what I know -- what Plotkin results?
  I always consider mentionning formal results a good thing}

@@

In this section however we show a more general approach to CBV CPS:
a CPS transformer. We take any implementation of |Symantics| (interpreter,
compiler, partial evaluator) and make it evaluate the CPS language
instead. We can view this transformation of interpreters as a
transformation on the object language, from direct to
continuation-passing style.

@@

\subsection{Lifting and Functor}
